Functors, CPS and monads, or how to generate efficient code from abstract designs

### 2005-04-07 paper/metamonads.tex
node changes

\usepackage{fancyvrb}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\}}
\newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
\setlength{\parskip}{0pt}
\newlength{\mathindent}
\setlength{\mathindent}{1em}

@@

\begin{small}
\begin{code}
let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
let param_code2 plus one =
  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
  and k0 x = x
  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
param_code2 plus one
\evalresult{.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.}
\end{code}
\end{small}

@@

\begin{small}
\begin{code}
module GenericArrayContainer(Dom:DOMAIN) =
  struct
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let get' x n m = .< (.~x).(.~n).(.~m) >.
  let get x n m = ret (get' x n m)
  let set' x n m y = .< (.~x).(.~n).(.~m) <- .~y >.
  let set x n m y = ret (set' x n m y)
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  let mapper g a = match g with
      | Some f -> .< Array.map (fun x -> Array.map .~f x) .~a >.
      | None   -> a
  let copy = (fun a -> .<Array.map (fun x -> Array.copy x) 
                       (Array.copy .~a) >. )
  (* this can be optimized with a swap_rows_from if it is known that
     everything before that is already = Dom.zero *)
  let swap_rows_stmt a r1 r2 =
      .< let t = (.~a).(.~r1) in
         begin 
             (.~a).(.~r1) <- (.~a).(.~r2);
             (.~a).(.~r2) <- t
         end >.
  let swap_cols_stmt a c1 c2 = .< 
      for r = 0 to .~(dim2 a)-1 do
          let t = (.~a).(r).(.~c1) in
          begin 
              (.~a).(r).(.~c1) <- (.~a).(r).(.~c2);
              (.~a).(r).(.~c2) <- t
          end
      done  >.
end
\end{code}
\end{small}
\noindent  Implementing other containers based on flattened vectors,
Fortran-style access, |Bigarray| or |Bigarray.Array2| is also possible.
Naturally, for proper use of |Bigarray| with specific types, it is 
important that the |DOMAIN| type matches -- but luckily the compiler 
will prevent us from making such a mistake.}

@@

\begin{small}
\begin{code}
module type OUTPUT = sig
  type contr  type res
  module D : DETERMINANT   module R : RANK
  module P : TRACKPIVOT
  val make_result : ('a,contr) code -> 
    (('a,res) code,
     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
       list, ('a,'w) code) monad
end
module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
    (Rank : RANK) = struct
  module Ctr = C(Dom)
  type contr = Ctr.contr
  type res = contr * Det.outdet * int
  module D = Det   module R = Rank
  module P = DiscardPivot
  let make_result b = mdo \{ det  <-- D.fin ();  rank <-- R.fin ();
    ret .< ( .~b, .~det, .~rank ) >. \}
end
\end{code}
\end{small}

@@

\begin{small}
\begin{code}
module type DETERMINANT = sig
  type indet  type outdet  type 'a lstate
  type tdet = outdet ref   
  val decl : unit -> 
    (unit, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val upd_sign : unit -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  ...
end
\end{code}
\end{small}
\noindent  Determinant tracking aspects involves being able to
generate code that defines variables used for tracking (|decl|),
generate code that updates the sign or the absolute
value of the determinant, and finally, converting the tracking state
to the final determinant value of the type |outdet|. GE of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| where |outdet| is |unit| and all the
functions of that module generate no code. For integer matrices, we
have to track some aspects of the determinant even if we don't output
it. The determinant tracking aspect is complex becasue tracking
variables, if any, are to be declared at the beginning of GE; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, which may be, for
example, a piece of code for the value of the type |float ref| among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. Other aspects, e.g., rank tracking, may use the
monadic state for passing of rank tracking variables. To be able to
compose determinant and rank tracking functors -- each of which may
(or may not) use the monadic state for passing its own data -- we make
an extensive use of open records (list of polymorphic variants
appeared to be the easiest way to implement such a union, in a pure
functional way). This lets us freely compose determinant-tracking,
rank-tracking, and other aspects.

Expressing domain constraints in functors (Div and integer domain).
Show the type error that occurs in improper instantiation.

Show several examples (e.g., code with NullPivot and nodet: no traces
of either det nor pivoting). Then the code with RowPivoting, Det and
float domain.

@@

>>>>

\usepackage{fancyvrb}
\DefineShortVerb{\|}
\DefineVerbatimEnvironment{code}{Verbatim}{xleftmargin=\mathindent,commandchars=\\\{\},fontsize=\small}
\newcommand{\evalresult}[1]{\ensuremath{\Longrightarrow}\textcolor{red}{#1}}
\setlength{\parskip}{0pt}
\newlength{\mathindent}
\setlength{\mathindent}{1em}

@@

\begin{code}
let letgen exp k = .<let t = .~exp in .~(k .<t>.)>.
let param_code2 plus one =
  let gen x y k = letgen (plus y one) (fun ce -> k (plus ce (plus x ce)))
  and k0 x = x
  in .<fun x y -> .~(gen .<x>. .<y>. k0)>.
param_code2 plus one
\evalresult{.<fun x_1 -> fun y_2 -> let t_3 = (y_2 + 1) in (t_3 + (x_1 + t_3))>.}
\end{code}

@@

\begin{code}
module GenericArrayContainer(Dom:DOMAIN) =
  struct
  type contr = Dom.v array array (* Array of rows *)
  type 'a vc = ('a,contr) code
  type 'a vo = ('a,Dom.v) code
  let get' x n m = .< (.~x).(.~n).(.~m) >.
  let get x n m = ret (get' x n m)
  let set' x n m y = .< (.~x).(.~n).(.~m) <- .~y >.
  let set x n m y = ret (set' x n m y)
  let dim2 x = .< Array.length .~x >.       (* number of rows *)
  let dim1 x = .< Array.length (.~x).(0) >. (* number of cols *)
  let mapper g a = match g with
      | Some f -> .< Array.map (fun x -> Array.map .~f x) .~a >.
      | None   -> a
  let copy = (fun a -> .<Array.map (fun x -> Array.copy x) 
                       (Array.copy .~a) >. )
  (* this can be optimized with a swap_rows_from if it is known that
     everything before that is already = Dom.zero *)
  let swap_rows_stmt a r1 r2 =
      .< let t = (.~a).(.~r1) in
         begin 
             (.~a).(.~r1) <- (.~a).(.~r2);
             (.~a).(.~r2) <- t
         end >.
  let swap_cols_stmt a c1 c2 = .< 
      for r = 0 to .~(dim2 a)-1 do
          let t = (.~a).(r).(.~c1) in
          begin 
              (.~a).(r).(.~c1) <- (.~a).(r).(.~c2);
              (.~a).(r).(.~c2) <- t
          end
      done  >.
end
\end{code}
\noindent  Implementing other containers based on flattened vectors,
Fortran-style access, |Bigarray| or |Bigarray.Array2| is also possible.
Naturally, for proper use of |Bigarray| with specific types, it is 
important that the |DOMAIN| type matches -- but luckily the compiler 
will prevent us from making such a mistake.}

@@

\begin{code}
module type OUTPUT = sig
  type contr  type res
  module D : DETERMINANT   module R : RANK
  module P : TRACKPIVOT
  val make_result : ('a,contr) code -> 
    (('a,res) code,
     [> `TDet of 'a D.lstate | `TRan of 'a R.lstate | `TPivot of 'a P.lstate]
       list, ('a,'w) code) monad
end
module OutDetRank(Dom:DOMAIN)(C: CONTAINER2D)
    (Det : DETERMINANT with type indet = Dom.v and type outdet = Dom.v)
    (Rank : RANK) = struct
  module Ctr = C(Dom)
  type contr = Ctr.contr
  type res = contr * Det.outdet * int
  module D = Det   module R = Rank
  module P = DiscardPivot
  let make_result b = mdo \{ det  <-- D.fin ();  rank <-- R.fin ();
    ret .< ( .~b, .~det, .~rank ) >. \}
end
\end{code}

@@

\begin{code}
module type DETERMINANT = sig
  type indet  type outdet  type 'a lstate
  type tdet = outdet ref   
  val decl : unit -> 
    (unit, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  val upd_sign : unit -> 
    (('a,unit) code, [> `TDet of 'a lstate ] list, ('a,'b) code) monad
  ...
end
\end{code}
\noindent  Determinant tracking aspects involves being able to
generate code that defines variables used for tracking (|decl|),
generate code that updates the sign or the absolute
value of the determinant, and finally, converting the tracking state
to the final determinant value of the type |outdet|. GE of a
floating-point matrix with no determinant tracking uses the
instantiation of |DETERMINANT| where |outdet| is |unit| and all the
functions of that module generate no code. For integer matrices, we
have to track some aspects of the determinant even if we don't output
it. The determinant tracking aspect is complex becasue tracking
variables, if any, are to be declared at the beginning of GE; the sign
of the determinant has to be updated on each row or column
permutation; the value of the determinant should be updated per each
pivoting. We use |lstate| to pass the tracking state, which may be, for
example, a piece of code for the value of the type |float ref| among
various determinant-tracking functions. The |lstate| is a part of the
overall monadic state. Other aspects, e.g., rank tracking, may use the
monadic state for passing of rank tracking variables. To be able to
compose determinant and rank tracking functors -- each of which may
(or may not) use the monadic state for passing its own data -- we make
an extensive use of open records (list of polymorphic variants
appeared to be the easiest way to implement such a union, in a pure
functional way). This lets us freely compose determinant-tracking,
rank-tracking, and other aspects.

The GE generator functor is as follows. It is
parameterized by the domain, container, pivoting policy (full, row,
nonzero, nopivoting), update policy (that use either ``fraction-less'
or full division), and by what to yield as the result. Some of the
argument modules such as |PIVOT| are functors themselves (parameterized
by the domain, the container, and the determinant functor). The sharing
constraints express obvious constraints on the instantiation of |Gen|,
for example, pivoting, determinant etc. components all use the same
domain. It must be stressed that all structures (i.e., module
instances) are stateless, and so we never have to worry that different
aspect functors (such as |CONTAINER2D| and |PIVOT|) are instantiated
with different but type-compatible instances of |DOMAIN|. That is, we
are not concerened at all about the value sharing. Aspects such as
determinant tracking may be stateful so that the determinant update
code have access to the determinant tracking variables declared
previsouly. But that state is handled via the monadic state. As we
have shown, open unions makes the overall monadic state compositional
with respect to the state of various aspects.

\begin{figure*}
\begin{code}
module Gen(Dom: DOMAIN)(C: CONTAINER2D)(PivotF: PIVOT)
          (Update: UPDATE with type baseobj = Dom.v and type ctr = C(Dom).contr)
          (Out: OUTPUT with type contr = C(Dom).contr and type D.indet = Dom.v 
                        and type 'a D.lstate = 'a Update.D.lstate) =
   struct
    module Ctr = C(Dom)
    module Pivot = PivotF(Dom)(C)(Out.D)
    type v = Dom.v
    let gen =
      let zerobelow b r c m n brc =
        let innerbody i = mdo \{
            bic <-- Ctr.get b i c;
            whenM (l1 LogicCode.not (LogicCode.equal bic Dom.zero ))
                (seqM (retLoopM (Idx.succ c) (Idx.pred m)
                          (fun k -> Update.update b r c i k) )
                      (Ctr.set b i c Dom.zero)) \} in 
        mdo \{
              seqM (retLoopM (Idx.succ r) (Idx.pred n) innerbody) 
                   (Update.update_det brc) \} in
      let dogen a = mdo \{
          r <-- Out.R.decl ();
          c <-- retN (liftRef Idx.zero);
          b <-- retN (Ctr.mapper Dom.normalizerf (Ctr.copy a));
          m <-- retN (Ctr.dim1 a);
          n <-- retN (Ctr.dim2 a);
          () <-- Update.D.decl ();
          () <-- Out.P.decl ();
          seqM 
            (retWhileM (LogicCode.and_ (Idx.less (liftGet c) m)
                                       (Idx.less (liftGet r) n) )
               ( mdo \{
               rr <-- retN (liftGet r);
               cc <-- retN (liftGet c);
               pivot <-- l1 retN (Pivot.findpivot b rr m cc n);
               seqM (retMatchM pivot (fun pv -> 
                        seqM (zerobelow b rr cc m n pv)
                             (Out.R.succ ()) )
                        (Update.D.zero_sign () ))
                    (Code.update c Idx.succ) \} ))
            (Out.make_result b) \} 
    in
    .<fun a -> .~(runM (dogen .<a>.)) >.
end
\end{code}
\end{figure*}

In addition to the ``regular'' type sharing constraints shown in the
|Gen| functor, there are also ``semantic'' sharing constraints, shown
in the following structure of the |UPDATE| signature:

\begin{code}
module DivisionUpdate
  (Dom:DOMAIN with type kind = domain_is_field)
  (C:CONTAINER2D)
  (Det:DETERMINANT with type indet=Dom.v) = struct ... end
\end{code}

The structure implements update policy that can use |Dom.div| operation without
restrictions. That is possible only if the domain has such an
operation. A domain such as integer domain may still provide |Dom.div|
of the same type, but that operation can only be used when we are sure
that the division is exact. Using the integer domain with
|DivisionUpdate| will lead to the compile-time error, when compiling
the \emph{generator} code. Thus, in some cases we can use module types
to express ``semantic'' constraints that cannot normally be expressed
via the types of module members.

\begin{code}
module GenFV4 = Gen(FloatDomain)
   (GenericVectorContainer)(FullPivot)
   (DivisionUpdate(FloatDomain)(GenericVectorContainer)(FDet))
   (OutDetRank(FloatDomain)(GenericVectorContainer)(FDet)(Rank))
module GenIA1 = Gen(IntegerDomain)
   (GenericArrayContainer)(RowPivot)
   (FractionFreeUpdate(IntegerDomain)(GenericArrayContainer)(IDet))
   (OutJustMatrix(IntegerDomain)(GenericArrayContainer)(IDet))
\end{code}

We can instantiate the |Gen| functor as shown above and inspect the
generated code, e.g., by printing |GenIA1.gen|. The code can then be
``compiled'' as |!. GenIA1.gen| or with off-shiring (cite). The code
for |GenFV4| shows full pivoting, determinant and rank tracking. The
code for all these aspects is fully inlined; no extra functions are
invoked and no tests other than those needed by the GE algorithm
itself are performed. The GE function returns a triple 
|float array array * float * int| of the U-factor, determinant and 
the rank. The
code generated by |GenIA1| shows absolutely no traces of determinant
tracking: no declaratrion of spurious variables, no extra tests,
etc. The code appears as if determinant tracking aspect did not exist
at all. The generated code for the above and other instantiations of
|Gen| can be examined at (cite).

@@

Combination of stateless modules (functors and structures) with
the monad with the compositional state greatly makes the aspects
freely composable without having to worry about presence or absense of
value aliasing. The only constraints to compositionality are the
typing constraints plus the constraints we specifically impose,
including semantic constraints (e.g., rings do not have the full
division).
